---
title: "Step Limit Analysis"
author: "Richard Lussier & Nic McPhee"
date: "6/5/2022"
output: 
  html_document:
    toc: true
    toc_depth: 3
params:
    data_file: "MutLIOZResults.txt"
    original_step_limit: 200
    graph_file_name: "temporary.png"
    test_problem: "Last Index of Zero"
    population_size: 500
    max_generations: 500
    num_training_cases: 200
    num_test_cases: 2000
    reproduction_operators: "50% Crossover, 50% UMAD (10% UMAD rate)"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library('ggplot2')
library('dplyr')
library('ggpubr')
#library('gridExtra')

dataFile <- params$data_file
originalStepLimit <- params$original_step_limit
graphFileName <- params$graph_file_name
testProblem <- params$test_problem
populationSize <- params$population_size
maxGenerations <- params$max_generations
numTrainingCases <- params$num_training_cases
numTestCases <- params$num_test_cases
reproductionOperators <- params$reproduction_operators
```

```{r functions, include=FALSE}

#This function combines all of the helper functions below. It takes the dataset, the step limit the problem was originally run on, and a string, which is the .png file you would like to export the graphs to (for example, "example.png").
GenerateGraphs <- function(givenData, originalStepLimit, saveToFile, groupPlots=TRUE) {
  
  givenData <- WrangleData(givenData)
  
  #Store the graphs into variables
  graph1 <- GraphOriginalGeneralizations(givenData, originalStepLimit)
  graph2 <- GraphOtherGeneralizations(givenData, originalStepLimit)
  graph3 <- GraphRunningOutOfSteps(givenData)
  
  #Put every graph into a list
  graphList <- list(graph1, graph2, graph3)
  
  #Print all graphs on one file if true, in separate files otherwise.
  if(groupPlots){
    allGraphs <- ggarrange(plotlist=graphList, ncol = 1, nrow = 3)
  
    ggexport(allGraphs, filename=saveToFile)
  } else {
    for(i in 1:length(graphList)){
      
      fileName <- paste(i, saveToFile, sep="_")
      ggexport(graphList[i], filename=fileName, width=800, height=800)
    }
  }
}

#This function takes a dataset with specific column names (RunNumber, Succeeded., Generalized., StepLimit, TotalError, NumFailedCases, and FailedByStepLimit) and wrangles them into the appropriate data type if needed. Adds a new column AllFailuresOutOfSteps
WrangleData <- function(givenDataFile) {
  # Read the data from the given file name
  givenData <- read.csv(givenDataFile, sep="")
  #Wrangle all of the data into their proper types in case they were imported in a weird format.
  givenData$RunNumber <- suppressWarnings(as.numeric(givenData$RunNumber)) 
  givenData$Succeeded. <- suppressWarnings(as.logical(givenData$Succeeded.))
  givenData$Generalized. <- suppressWarnings(as.logical(givenData$Generalized.))
  givenData$StepLimit <- suppressWarnings(as.numeric(givenData$StepLimit))
  givenData$TotalError <- suppressWarnings(as.numeric(givenData$TotalError))
  givenData$NumFailedCases <- suppressWarnings(as.numeric(givenData$NumFailedCases))
  givenData$FailedByStepLimit <- suppressWarnings(as.numeric(givenData$FailedByStepLimit))
  
  #Add a new column, a boolean that says whether all of the failed cases came from running out of steps.
  givenData$AllFailuresOutOfSteps = (givenData$NumFailedCases == givenData$FailedByStepLimit)

  #Return the wrangled dataset
  return(givenData)
}

#This function takes a dataset and a step limit and returns a graph of every run that generalized at the given step limit.
GraphOriginalGeneralizations <- function(givenData, originalStep) {
  
  #Filter the runs to only the ones that succeeded at the given step limit. Get a vector of only the run numbers.
  runsThatSucceeded <- subset(givenData, (givenData$StepLimit == originalStep & givenData$Succeeded.))$RunNumber

  #Filter the runs to only the ones that generalized at the given step limit. Get a vector of only the run numbers.
  runsThatGeneralized <- subset(givenData, (givenData$StepLimit == originalStep & givenData$TotalError == 0))$RunNumber
  
  #Filter from the original dataset to only grab the ones that generalized at the given step limit
  fullGenData <- givenData %>%
  filter(RunNumber %in% runsThatGeneralized)
  
  #Create the graph, organized and colored by run number
  graph <- ggplot(fullGenData, aes(x=StepLimit, y=NumFailedCases, group=factor(RunNumber), color=factor(RunNumber))) + geom_line(alpha=0.5, show.legend=FALSE) + ggtitle(paste("Runs that generalized at step-limit", originalStep, sep=" "))
  
  #Return the graph
  return(list(graph=graph, numSucceeded=length(runsThatSucceeded), numGeneralized=length(runsThatGeneralized)))
}

#This function takes the dataset and the step limit the problem was originally run on and returns a graph of every run that generalizes not at the given step limit.
GraphOtherGeneralizations <- function(givenData, originalStep) {
  
  #Grab all of the runs that generalized at some point
  allGeneralizations <- givenData %>%
  group_by(RunNumber) %>%
  filter(min(NumFailedCases) == 0)
  
  #Filter the runs to only the ones that generalized anywhere thats not the original step limit given and grab the run number
  generalizedElsewhere <- subset(allGeneralizations, (StepLimit == originalStep & TotalError > 0))$RunNumber
  # print(paste("The number of runs that generalized with step limit !=", originalStep, "was", length(generalizedElsewhere), sep=" "))
  
  #Filter from the original dataset to only grab the ones that generalized somewhere other than the given step limit
  fullRuns <- givenData %>%
  filter(RunNumber %in% generalizedElsewhere)
  
  #Graph the runs, colored by their group number.
  graph <- ggplot(fullRuns, aes(x=StepLimit, y=NumFailedCases, group=factor(RunNumber), color=factor(RunNumber))) + geom_line(alpha=0.5, show.legend=FALSE)  + ggtitle(paste("Runs that generalized somewhere not step-limit", originalStep, sep=" "))
  
  return(list(graph=graph, numGeneralized=length(generalizedElsewhere)))
}

#This function takes a dataset and a number which corresponds to the last step limit of the dataset. This function returns a graph containing every run that was still running out of steps when the last step limit was computed.
GraphRunningOutOfSteps <- function(givenData) {
  
  #Get the highest step limit
  lastStepLimit <- max(givenData$StepLimit)
  
  #Take the maximum number of failed cases throughout the entire dataset (hoping one of them is close to the number of total cases) and multiply it by an arbitrary number so we can filter runs and make graphs cleaner
  failedCasesThreshold <- max(givenData$NumFailedCases) * 0.05
  
  #Gets all runs that are below the given threshold and didn't generalize (min != 0)
  fewFailedCases <- givenData %>%
  group_by(RunNumber) %>%
  filter(min(NumFailedCases) < failedCasesThreshold & min(NumFailedCases) != 0)
  
  #Gets the run number from the above dataset
  outOfSteps <- subset(fewFailedCases, (StepLimit == lastStepLimit) & AllFailuresOutOfSteps)$RunNumber
  
  #get the runs that ran out of steps at the max step limit
  getRuns <- givenData %>%
  filter(RunNumber %in% outOfSteps)
  
  graph <- ggplot(getRuns, aes(x=StepLimit, y=NumFailedCases, group=factor(RunNumber), color=factor(RunNumber))) + geom_line(alpha=0.5, show.legend=FALSE)  + ggtitle(paste("Runs that are running out of steps at", lastStepLimit, sep=" "))
  
  return(list(graph=graph, numRunningOut=length(outOfSteps)))
}
```

```{r include=FALSE}
dataset <- read.csv(params$data_file, sep="")

GenerateGraphs(params$data_file, originalStepLimit, graphFileName, groupPlots=FALSE)
```

## Overview & parameters

These results were for `r testProblem` with the following parameters:

* Original step limit = `r originalStepLimit`
  * Min step limit = `r min(dataset$StepLimit)`
  * Max step limit = `r max(dataset$StepLimit)`
* Population size = `r populationSize`
* Maximum generations = `r maxGenerations`
* Number of training cases = `r numTrainingCases`
* Number of test cases = `r numTestCases`
* Reproduction operators = `r reproductionOperators`

The results were in the "`r dataFile`" data file. Files for the
corresponding graphs can be found at files with the names "*_`r graphFileName`"
where "*" is replaced by 1, 2, and 3.

## Runs that generalized at step limit `r originalStepLimit`

```{r echo=FALSE}
givenData <- WrangleData(params$data_file)

originalGeneralizesResults = GraphOriginalGeneralizations(givenData, originalStepLimit)
```

The graph below shows the number of failed cases for various step limits for all
the runs that generalized at the original step limit, i.e., `r originalStepLimit`.
There were `r originalGeneralizesResults$numSucceeded` runs that succeeded and `r originalGeneralizesResults$numGeneralized` runs that generalized at step limit
`r originalStepLimit`.
All of these runs should have zero failed cases at step limit 
`r originalStepLimit`, but some might fail one or more cases at higher or
lower step limits.

```{r echo=FALSE}
originalGeneralizesResults$graph
```

## Runs that generalized elsewhere than step limit `r originalStepLimit`

```{r echo=FALSE}
otherGeneralizations = GraphOtherGeneralizations(givenData, originalStepLimit)
```

The graph below shows the number of failed cases for various step limits for all
the runs that _didn't_ generalize at the original step limit (i.e., `r originalStepLimit`),
but _did_ generalize at at least one other step limit. There were `r otherGeneralizations$numGeneralized` such runs.

```{r echo=FALSE}
otherGeneralizations$graph
```

## Runs that failed because they still ran out of steps at step limit `r max(dataset$StepLimit)`

```{r echo=FALSE}
outOfSteps = GraphRunningOutOfSteps(givenData)
```

The graph below shows the number of failed cases for various step limits for all
the runs that failed all test cases because they ran out of steps. There were
`r outOfSteps$numRunningOut` such runs.

```{r echo=FALSE}
outOfSteps$graph
```
